Stickleback EDA

### 2023-04-04 SticklesEDA.Rmd
updates including plots

#Loading Libraries
```{r}
library(tidyverse)
library(ggplot2)
library(readr)
library(dplyr)
library(DT)

@@

```{r}
stickle <- read_csv('stickleback_fossil_vs_extant.csv')
summary(stickle)
#seems every column outside of ID has missing data to some extent
#Highest amount of NA's is cle.sc column
#least amount of NAs in mds column
```

```{r}
datatable(stickle)

#Should I split up univID?
#Maybe, maybe not
```
```{r}
stickles <- datatable(stickle)
grepl("M",stickle$univID, fixed=TRUE)
stickle$univID
stick <- stickle %>% 
  mutate(gender = ifelse(grepl("M",univID,fixed=TRUE),"M","F"))

table(stick$gender)
#Only 130 Males?????
#sheesh

@@

>>>>

#Loading Libraries

```{r}
library(tidyverse)
library(janitor)
library(tidymodels)
library(furrr)
library(doParallel)

@@

```{r}
stickle <- read_csv('stickle_revamped.csv')
ggplot2(x=as.integer(stick %>% filter(!is.na(gender)) %>% pull(gender)),stick %>% filter(!(is.na(gender))) %>% pull(ips.prop))

gens <- stick %>% filter(!is.na(gender)) %>% pull(gender)
props <- stick %>% filter(!(is.na(gender))) %>% pull(ips.prop)
ggplot(data=stick, aes(x = gens,y=props)) + geom_bar()
summary(stickle)
#seems every column outside of ID has missing data to some extent
#Highest amount of NA's is cle.sc column
#least amount of NAs in mds column
```

```{r}
#generating gender for the dataset

stick <- stickle %>% 
  mutate(gender = as.factor(case_when(
    grepl(".M.",univID,fixed=TRUE) ~ "Male",
    grepl(".F.",univID,fixed=TRUE) ~ "Female",
    grepl("Fos",univID,fixed=TRUE) ~ NA_character_
    )), ips = as.factor(ips)
  ) %>% 
  select(gender, everything())

stick_lakes_bl <- stick %>% 
  mutate( B_L = case_when( 
    grepl("Emi.",univID,fixed=TRUE) ~ "Emily Lake",
    grepl("LQx.",univID,fixed=TRUE) ~ "Little Quarry Lake",
    grepl("Pax.",univID,fixed=TRUE) ~ "Pax Lake",
    grepl("Pri.",univID,fixed=TRUE) ~ "Pri Lake",
    grepl("Joe.",univID,fixed=TRUE) ~ "Joe Lake",
    grepl("Nor.",univID,fixed=TRUE) ~ "Nor Lake",
    grepl("Swa.",univID,fixed=TRUE) ~ "Swa Lake",
    grepl("The.",univID,fixed=TRUE) ~ "Theimer Lake",
    grepl("Vib.",univID,fixed=TRUE) ~ "Village Bay Lake",
    grepl("Fos.",univID,fixed=TRUE) ~ "Fossil"
  ), Ecotype = case_when(
    grepl(".B.",univID,fixed=TRUE) ~ "Benthic",
    grepl(".G.",univID,fixed=TRUE) ~ "Generalist",
    grepl(".L.",univID,fixed=TRUE) ~ "Limnetic",
    grepl(".K.",univID,fixed=TRUE) ~ NA_character_

  )

  )

stick %>% summary(ips)
range(stick$ips)
stick %>% filter(is.na(ect))

#ect = bottom of fish, belly

@@

```{r}

extant_stickle <- stick %>% filter(gender != "Fossil")

Visualizing the missing variables found within dataset.

```{r}
# install.packages('visdat')
# install.packages('UpSetR')
# install.packages('naniar')
library(naniar)
library(visdat)
library(UpSetR)
```

how many missing obs per variable

```{r}
gg_miss_var(stick_lakes_bl)
vis_miss(stick_lakes_bl)
```

```{r}
gg_miss_var(stick_no_gender)
vis_miss(stick_no_gender)

Are they often missing in conjunction with one another?

```{r}
n_var_miss(stick) # gives all variables with missing data
gg_miss_upset(stick) # gives plot that captures combination of missingness in data
gg_miss_upset(stick,nsets = n_var_miss(stick))
gg_miss_upset(stick,nsets=2) # gives plot that captures combination of missingness in data

#cle - shoulder bone buried or not there in fossils.
# NA only, needs a cle for a functional fish

#jawbon - pmx. is always there for a functional fish.

How about when excluding gender from the dataset?

```{r}

stick_no_gender <- stick %>% 
  select(-gender, -univID)
gg_miss_upset(stick_no_gender)
gg_miss_upset(stick_no_gender,nsets = n_var_miss(stick_no_gender))

```{r}
ggplot(stick,
       aes(x = cle.sc,
           y = cle)) +
 geom_miss_point() + 
 facet_wrap(~gender)

Might be concerning. Very prelim look, but first thoughts are that the fossil group seems to have cle.sc values between 5-6 that are missing, rest are not. more too look into later.

```{r}
stick_no_id <- stick %>% select(!univID)
plot_bar(imputation.list = stick_imputed, var.name = "gender",
    original.data = stick_no_id)
```

```{r}
plot_box(imputation.list = stick_imputed, var.name = "ect",
    original.data = stick_no_id)
```

```{r}

```{r}
# devtools::install_github("agnesdeng/mixgb") #installs the mixgb package which utilizes xgboost as imputation method for MI
stick_clean <- stick %>% 
  select(!univID)

stick_clean <- data_clean(stick_clean)
```

```{r}
#Finding the best rounds, very basic CV. Still need to tune other params outside of gender. confused about the best rounds

library(mixgb)

cv.results <- mixgb_cv(data = stick_clean, verbose = FALSE)

cv.results$response

cv.results$best.nrounds

cv.results2 <- mixgb_cv(data = stick_clean, nfold = 5, nrounds = 100,
    early_stopping_rounds = 1, response = "gender", verbose = FALSE)

cv.results2$best.nrounds
```

```{r}
#trying to tune parameters, bit lost at this

best_rounds <- c()
for(i in 1:50){
cv.results_ <- mixgb_cv(data = non_gender, verbose=FALSE,nrounds=200)
best_rounds <-append(best_rounds, cv.results_$best.nrounds)
if(best_rounds[i]==max(best_rounds)){
  maxresponse = cv.results_$response
  maxRounds <- cv.results_$best.nrounds
}

}

max(best_rounds)
maxresponse

cv.results_ <- mixgb_cv(data = non_gender,response="stl", verbose=FALSE,nrounds=200)
cv.results_$best.nrounds

parallel::detectCores()
n.cores <- parallel::detectCores() - 1

my.cluster <- parallel::makeCluster(
  n.cores, 
  type = "FORK"
  )
doParallel::registerDoParallel(cl = my.cluster)
foreach::getDoParRegistered()
foreach::getDoParWorkers()

best_rounds <- c()
x <- foreach(
  i = 1:1000, 
  .combine = 'c'
) %dopar% {
  cv.results_ <- mixgb_cv(data = stick_clean, verbose=FALSE,nrounds=200)
  best_rounds <-append(best_rounds, cv.results_$best.nrounds)
}  
best_rounds
max(x)

```{r}
#imputed dataset
stick_imputed <- mixgb(data = stick_clean, m = 15, nrounds = 30)

# imputed_data_cvresp_else <- mixgb(data = stick_clean, m = 20, nrounds = 75)
# stick_imputed_e <- imputed_data_cvresp_else

```{r}
#making a function to grab means cause I'm lazy

grab_means <- function(data, i){
  data[[i]] %>% 
    summarize(across(where(is.numeric),mean))
  }

#grab_means(stick_imputed,1)

imputed_means <- c(1:10) %>% 
  future_map_dfr(grab_means,data=stick_imputed)

imputed_means

```{r}

#Comparing results to complete case dataset and non completed, non imputed values

mean_extant_complete <- stick %>% 
  na.omit %>% 
  summarise(across(where(is.numeric),mean))
mean_extant_complete
```

```{r}
all_cores <- parallel::detectCores(logical = FALSE)
library(doParallel)

cl <- makePSOCKcluster(all_cores)
registerDoParallel(cl)
```

```{r}

set.seed(100)
extant <- stick %>% 
  drop_na(gender) %>%
  select(-c(univID, gender))
extant_gender <- stick %>% 
  drop_na(gender) %>% 
  pull(gender)

cv.results_ <- mixgb_cv(data = extant, verbose=FALSE, nfold=10,nrounds=100)
cv.results_$best.nrounds
cv.results_$response

extant_imputed <- mixgb(data = extant, m = 10, nrounds = 100)

extant %>% mutate(gender = c(extant_gender)) %>% select(gender, everything())
add_gender <- function(data,i){
  data[[i]] %>% 
    mutate(gender = c(extant_gender)) %>% select(gender, everything())

}
imp_extant_with_gender <- c(1:10) %>% 
  map(data=extant_imputed,add_gender)

```{r}
library(tidyverse)
library(tidymodels)
library(rsample)

```{r}

#general accuracy with imputed datasets(only including extant)
#maybe figure out a way to combine the results? idk

getAccuracy  <- function(data,i){

  split <- initial_split(data[[i]], prop = 0.7, strata = gender)
  train <- training(split)
  test <- testing(split)

  train_fold <- vfold_cv(train, v = 10, strata = gender)

  xg_spec <- boost_tree(trees = tune(),
                        tree_depth = tune(), 
                        min_n = tune(), 
                        loss_reduction = tune(),
                        sample_size = tune(), 
                        mtry = tune(),
                        learn_rate = tune()) %>% 
    set_engine("xgboost") %>% 
    set_mode("classification")

  xg_rec <- recipe(gender ~., data = train) %>% 
    step_dummy(all_nominal_predictors())

  xg_wf <- workflow() %>% 
    add_model(xg_spec) %>% 
    add_recipe(xg_rec)
  xg_grid <- grid_latin_hypercube(trees(),
                                  tree_depth(),
                                  min_n(),
                                  loss_reduction(),
                                  sample_size = sample_prop(),
                                  finalize(mtry(), train),
                                  learn_rate(),
                                  size = 100)
  xg_tune <- tune_grid(xg_wf,
                       resamples = train_fold,
                       grid = xg_grid,
                       control = control_grid(save_pred = TRUE))

  best_acc <- xg_tune %>% 
    select_best("accuracy")

  xg_final <- finalize_workflow(
    xg_wf,
    best_acc
  )
  xg_last <- last_fit(xg_final, split)

   data_output <- xg_last
  #   collect_predictions() %>% 
  #   accuracy(gender, .pred_class)

  return(data_output)

}

imp_extant_with_gender[[1]]
```

```{r}
library(furrr)

# list <- c(1:10) %>% 
#   map(getAccuracy, data=imp_extant_with_gender)
list_of_acc <- list()
for(i in 1:10){
  list_of_acc[[i]] <- getAccuracy(data=imp_extant_with_gender, i)
}

list2 <- list()
list2[[1]] <- getAccuracy(data=imp_extant_with_gender,1)

imputed_data_tuned <- mixgb(data = non_gender, m=5, nrounds = 140)
c(1:10)

getacc <- function(data,i){
  data[[i]] %>% 
  collect_predictions() %>% accuracy(gender,.pred_class)
}
accuracy_per_imputed_data <- c(1:10) %>% 
  map_dfr(getacc,data=list)
accuracy_per_imputed_data
```

```{r}
library(mice)
# install.packages("mitools")
# install.packages('miceadds')
library(mitools)
library(miceadds)
with(stick_imputed, mean)
as.mids(stick_imputed)
help(with)
stick_imputed
stick %>% 
  grepl()
stick_imputed
all_numeric <- function(data,i){ 
  data[[i]] %>% 
    select(!c("gender","ips"))
}

test4 <- with(test3,expr=mean)
test[[1]]
stick_test <- imputationList(stick_imputed)
#getting time variable
univ <- stick %>% 
  mutate(time = as.integer(sub(".*([0-9]{2})\\..*", "\\1",univID)))

times <- univ %>% drop_na(time) %>% pull(time)

times_vector <- c(rep(1,43)) %>% 
  append(21-times)

means<- function(data){
  data %>% 
    summarise(across(where(is.numeric),mean))
}
abc <- with(stick_test, fun=means)
MIcombine(abc)
mm<- as.mira(abc)

summary(MIcombine(abc))

add_time<- function(data){
data %>% 
  slice(368:1181) %>%
  mutate(time = times_vector) %>% 
  select(time, everything())
}

means_by_time <- function(data){
  data %>% group_by(time, gender) %>% 
  summarize(across(where(is.numeric),mean))
}

```{r}
mean_diff <- function(data){
  data %>% group_by(time, gender) %>% 
  summarize(across(where(is.numeric),mean)) %>% 
  summarize(across(where(is.numeric),diff))

}

mean_diff2 <- function(data){
  data %>% 
  summarize(across(where(is.numeric),diff))

}

stick_time[[3]] %>% group_by(time,gender) %>% summarize(across(where(is.numeric),mean))

#this is an issue
#basically, sometimes the imputation doesn't have values for both sexes at a given time point

stick_imputed <- mixgb(data = stick_clean, m = 10, nrounds = 15)

stick_time <- with(imputationList(stick_imputed), fun=add_time)

stick_means <- with(imputationList(stick_time), fun=means_by_time)
mean_differences <- with(imputationList(stick_means[-c(2,4,8,9)]),fun=mean_diff2)
mean_differences

```{r}
fin <- (mean_differences[[1]]+mean_differences[[2]]+mean_differences[[3]]+mean_differences[[4]]+mean_differences[[5]]+mean_differences[[6]])/6

```{r}
fin %>% ungroup() %>%
  select(where(is.numeric)&!time) %>% colnames() %>% 
  map( ~ {ggplot(fin, aes(x = time,y=.data[[.]])) + geom_line()})

Mean difference has gone up by a decent bit over the course of t=0 to t=21

```{r}
fin %>% ungroup() %>%
  select(where(is.numeric)&!time) %>% colnames() %>% 
  map( ~ {ggplot(fin, aes(x = time,y=.data[[.]])) + geom_line()})
